---
title: "Exercise #3"
subtitle: "Fortgeschrittene Statistische Software für NF"
author: "Anastasiia Petrus (12521274), Nils Przybylla (12734986)"
date: "`r Sys.Date()`"
output: distill::distill_article
editor_options: 
  markdown: 
    wrap: 72
---

## Exercise 1: Initializing git (4 Points)

For this whole exercise sheet we will be tracking all our changes to it
in git.

a)  Start by initializing a new R project with git support, called
    `2025-exeRcise-sheet-3`. If you forgot how to do this, you can
    follow this
    [guide](https://malikaihle.github.io/Introduction-RStudio-Git-GitHub/rstudio_project.html).
b)  Commit the files generated by Rstudio.
c)  For all of the following tasks in this exercise sheet we ask you to
    always commit your changes after finishing each subtask e.g. create
    a commit after task *1d*, *1e* etc.

> Note: This applies only to answers that have text or code as their
> answer. If you complete tasks in a different order or forget to commit
> one, this is no problem. If you change your answers you can just
> create multiple commits to track the changes.

d)  Name 2 strengths and 2 weaknesses of git. (Don't forget to create a
    commit after this answer, see *1c*)

    **Strengths:**

    1.  History allows you to go back and understand changes or revert
        when there are problems
    2.  Great for collaboration

    **Weaknesses:**

    1.  Can be a bit complicated to use
    2.  Struggles with binary (i.e. non-text) files

e)  Knit this exercise sheet. Some new files will automatically be
    generated when knitting the sheet e.g. the HTML page. Ignore these
    files, as we only want to track the source files themselves. You
    can, but don't need to create a `.gitignore` file. Just do not
    commit these files manually.

## Exercise 2: Putting your Repository on GitHub (3 Points)

For this task you will upload your solution to GitHub.

a)  Create a new repository on GitHub in your account named
    `exeRcise-sheet-3`. Make sure you create a **public repository** so
    we are able to see it for grading. Add the link to the repository
    below:
https://github.com/nilz1989/exeRcise-sheet-3.git
b)  Push your code to this new repository by copying and executing the
    snippet on github listed under
    `…or push an existing repository from the command line`.
c)  Regularly push your latest changes to GitHub again and especially do
    so when you are finished with this sheet.

## Exercise 3: Pixar Films (4 Points)

Download the `pixar_films` and `public_response` datasets from the
GitHub repository and track them in git.

Link:
<https://github.com/rfordatascience/tidytuesday/tree/main/data/2025/2025-03-11>

For small datasets like these adding them to git is not a problem.

a)  Load the `pixar_films` dataset into R. Clean the dataset by removing
    films without a title. Inspect the variable `film_rating`. What are
    the possible values and what do they mean? Create a factor variable
    for the film rating. Why is this appropriate?

```{r}
library(readr)
library(dplyr)

pixar_films <- read_csv("pixar_films.csv")
```

```{r}
pixar_films_clean <- pixar_films %>% filter(!is.na(film) & film != "")
print(head(pixar_films_clean, 10))
```

From the dataset, we observe the following values:

"G": General audiences — all ages admitted.

"PG": Parental Guidance Suggested — some material may not be suitable
for children.

"N/A" or NA: Missing or unavailable rating (e.g., for very recent or
unreleased films like Lightyear or Turning Red).

```{r}
pixar_films_clean <- pixar_films_clean %>%
  mutate(film_rating_factor = factor(film_rating))

pixar_films_new <- pixar_films_clean %>%
  select(-film_rating)
```

Creating a factor is appropriate because:

1.  Categorical nature: Film ratings are categories (e.g., "G", "PG")
    and not continuous or numeric values.

2.  Useful for analysis: Factors are treated properly in statistical
    models (e.g. regression) or in group-wise summaries (like counting
    how many films per rating)

3.  Ensures ordering if needed: We can specify levels to reflect an
    order (e.g., "G" \< "PG"), if appropriate for some tasks.

<!-- -->

b)  Inspect the film titles manually. Which films form a film series? A
    film series can be identified by a common word in the titles of the
    films, often in conjunction with a number in the title,
    e.g. "Despicable Me" and "Despicable Me 2". Create a dataframe which
    displays a list of the different series with the titles of the films
    and how many films belong to the series. Output the dataframe.

```{r}
library(stringr)
library(dplyr)

film_series_alternative <- pixar_films_clean %>%
  mutate(series = case_when(
    (str_detect(film, regex("Toy|Story"))) ~ "Toy Story",
    (str_detect(film, regex("Cars"))) ~ "Cars",
    (str_detect(film, regex("Monsters"))) ~ "Monsters",
    (str_detect(film, regex("Incredibles"))) ~ "Incredibles",
    (str_detect(film, regex("Finding|Nemo|Dory"))) ~ "Finding",
    
    TRUE ~ film
  ))

film_series_summary_alternative <- film_series_alternative %>%
  group_by(series) %>%
  summarise(
    titles = paste(film, collapse = ", "),
    count = n(),
    .groups = "drop"
  ) %>% filter(count >= 2)

print(head(film_series_summary_alternative, 10))
```

c)  Load the `public_response` dataframe into R. Convert the
    `cinema_score` variable into a factor while ensuring the factor
    levels are defined in ascending order, from the lowest to the
    highest score. Combine `public_response` with the `pixar_films`
    dataset using an appropriate merge variable.

```{r}
library(readr)
library(dplyr)

public_response <- read_csv("public_response.csv")
```

```{r}
public_response$cinema_score <- factor(
  public_response$cinema_score,
  levels = c("A–", "A", "A+"),
  ordered = TRUE
)
```

```{r}
library(dplyr)

pixar_combined <- left_join(pixar_films_new, public_response, by = "film")
print(head(pixar_combined, 10))
```

d)  Choose one of the variables representing the public response and
    create a bar plot for the films belonging to a series. Here are the
    details of the plot:

    -   The film series are represented on the x-axis.
    -   Your chosen public response variable is displayed on the y-axis.
    -   Each film in the series is represented as a separate bar. Bars
        are grouped by film under their respective series on the x-axis.
        Order the bars within a series according to the release date of
        the films.
    -   A title and axis labels for context.

    What do you notice when comparing the scores of the films in a
    series? Do you see any patterns?

We’ll use rotten_tomatoes because it provides a numeric score for the
majority of films.

install.packages("viridis")

```{r}
filtered_series_data <- film_series_alternative %>%
  semi_join(film_series_summary_alternative, by = "series")

series_scores_ordered_alternative <- pixar_combined %>%
  inner_join(filtered_series_data %>% 
  select(film, series), by = "film") %>%
  arrange(series, release_date) %>%
  mutate(film = factor(film, levels = unique(film)))

ggplot(series_scores_ordered_alternative, aes(
  x = series,
  y = rotten_tomatoes,
  fill = film
)) +
  geom_bar(
    stat = "identity",
    position = position_dodge(),
    color = "black"
  ) +
  labs(
    title = "Rotten Tomatoes Scores by Film Series",
    x = "Film Series",
    y = "Rotten Tomatoes Score",
    fill = "Film"
  ) +
  scale_fill_viridis_d(option = "C") +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Cars shows a decline in reception (Cars 2 scored especially low).

Finding Nemo vs Finding Dory – both are well received, but original
tends to score slightly higher.

Incredibles has strong and consistent ratings across both films.

Toy Story scores remain consistently high across the series (often
97–100).

Many Pixar originals tend to outperform sequels (like the Monsters
movies), except for franchises like Toy Story that maintain quality.

## Exercise 4: Open Analysis (4 points)

This exercise is a bit more open-ended. You can choose any dataset from
[Our World in Data](https://ourworldindata.org/) and analyze it, while
determining the research question yourself.

a)  Go to <https://github.com/owid/owid-datasets/tree/master/datasets>
    and choose a dataset that interests you. You can have a look at
    <https://ourworldindata.org/> to gather some inspiration.
b)  Download the dataset and track it in git.
c)  Put the name / title of the dataset and a link to it below.

-   Dataset Name: **Food waste in the European Union - Europa (2010)**
-   Link:
    [https://github.com/owid/owid-datasets/tree/master/datasets/Food%20waste%20in%20the%20European%20Union%20-%20Europa%20(2010)#food-waste-in-the-european-union---europa-2010](https://github.com/owid/owid-datasets/tree/master/datasets/Food%20waste%20in%20the%20European%20Union%20-%20Europa%20(2010)#food-waste-in-the-european-union---europa-2010){.uri}

d)  Come up with a (research) question you want to answer with the data
    and briefly explain why you believe this is an interesting question
    within one sentence. It should be a question that can be answered
    with the dataset and using R.

Which European countries rank highest in food waste by sector
(Manufacturing, Households, Other), and how do these rankings differ
across sectors?"

Why this question is interesting:

It allows us to identify whether specific sectors are more responsible
for food waste in different countries.

Sector-specific insights can help guide targeted policy interventions
(e.g., encouraging more sustainable household behavior or improving
manufacturing efficiency).

Comparing sectoral rankings can reveal structural or cultural
differences in food systems between countries.

e)  Use R to answer your chosen question. Please limit your analysis to
    the functions and techniques we have covered so far in the course.
    You are **not expected** to use advanced statistical models or
    external packages which haven't been introduced.

```{r}
library(tidyverse)

food_waste <- read_csv("Food_waste_in_the_European_Union_2010.csv")

food_waste_clean <- food_waste %>%
  filter(Entity != "European Union")

food_waste_long <- food_waste_clean %>%
  pivot_longer(cols = c(`Manufacturing (Europa (2010))`, 
                        `Households (Europa (2010))`, 
                        `Other (Europa (2010))`),
               names_to = "Sector", values_to = "Food_waste_tonnes")

food_waste_long <- food_waste_long %>%
  mutate(Sector = case_when(
    str_detect(Sector, "Manufacturing") ~ "Manufacturing",
    str_detect(Sector, "Households") ~ "Households",
    str_detect(Sector, "Other") ~ "Other"
  ))

top10_total <- food_waste_long %>% 
  group_by(Entity) %>%
  summarise(total_waste = sum(Food_waste_tonnes, na.rm = TRUE)) %>%
  slice_max(total_waste, n=10) %>%
  pull(Entity)

top10_overall_df <- food_waste_long %>%
  filter(Entity %in% top10_total)
```

f)  Create a meaningful plot / figure with the dataset. Make sure to
    provide a figure caption (via the chunk options / Rmarkdown) and
    correctly label the figure.

```{r}
library(scales)

ggplot(top10_overall_df, aes(x = reorder(Entity, -Food_waste_tonnes), 
                              y = Food_waste_tonnes, fill = Sector)) +
  geom_col(position = "dodge", color = "black") +
  scale_y_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  labs(title = "Top 10 EU Countries by Food Waste per Sector (2010)",
       x = "Country",
       y = "Food Waste (tonnes)",
       fill = "Sector") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)
)
```

General observations:

The United Kingdom is the top contributor to total food waste, with particularly high waist levels in the household sector.

Germany and France also report substantial food waste, but more balance distribution across all three sectors. 

Countries like Poland, Italy, and the Netherlands show especially high waisted in the manufacturing sector. 

Spain stands out for having a relatively large contribution from the other category, compared to most other countries.

Belgium, Romania, and Sweden round out the top 10 with overall lower waist levels, but still notable.

Sectoral trends: 

Households are the largest source of food waste across most countries, particularly in the UK, Germany, and France. 

The manufacturing sector rank second overall, but leads in countries like Poland, the Netherlands and Italy.

The other category typically contributes the least though countries like Spain and the UK show notable values here as well.

Key takeaways:

Household behavior plays a dominant role in food waste generation across Europe

The industrial footprint of countries influences the proportion of waste from manufacturing.

Sector-specific differences suggest the need for tailored food, waste reduction strategies for each country.

## Final Note

Make sure to push all your commits and changes to GitHub before
submitting the exercise sheet.
